---
title: "1 Solutions - R Fundamentals"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this document we will briefly practice working with different vector types.  Follow the instructions in the comments of each code chunk.

## Exercise 1 - Creating and examining vectors

```{r}
# In a recent coding competition, Liz came first, Alex came second 
# and Jiena came third.

# here is a vector with their names
name <- c("Alex", "Jiena", "Liz")

# Check that this vector is a character vector of length 3
str(name)
```

```{r}
# Create a corresponding numeric vector with each person's position.
position <- c(2, 3, 1)
```

```{r}
# Check that your vector is a numeric vector of length three.
str(position)
```

## Exercise 2 - Converting vectors to factors and ordered factors 

```{r}
# Convert the 'name' character vector to a factor vector called 'name_factor'.
# Check that the conversion succeeded.
name_factor <- as.factor(name)
str(name_factor)
```


```{r}
# Convert your 'name' character vector to an ordered factor vector
# in increasing position order. Check that the conversion succeeded. 
name_ordered <- ordered(name, levels = c("Jiena", "Alex", "Liz"))
str(name_ordered)
```

```{r}
# EXTENSION: Use a function to display the vector in order of position (highest to lowest)?
sort(name_ordered, decreasing = TRUE)
```

## Exercise 3 - Loading and inspecting dataframes

```{r}
# Here is a URL for on online csv of data about charity donations
url <- "https://peopleanalytics-regression-book.org/data/charity_donation.csv"

# Download this into a dataframe with a name of your choice
charity_data <- read.csv(url)

```

```{r}
# Inspect the dataframe to find out how many rows of data there are
# and what the columns names and data types are
str(charity_data)

```

## Exercise 4 - Using functions, packages and the pipe operator

```{r}
# Load the dplyr package 
library(dplyr)

# Seek help on the functions filter() and pull() from dplyr
# What do these functions do and what are their arguments?
?dplyr::filter
?dplyr::pull

```

```{r}
# Using the pipe operator, write code to calculate the average amount
# donated by Rural Domestic individuals, rounded to the nearest dollar.
# Remember that == (double equals) is used for precise equality
charity_data |> 
  dplyr::filter(reside == "Rural Domestic") |> 
  dplyr::pull(total_donations) |> 
  mean() |> 
  round()
```

```{r}
## EXTENSION: Write similar code to calculate the average amount donated 
## by males the age range of 30-39 inclusive, rounded to the nearest dollar
charity_data |> 
  dplyr::filter(gender == "M", age %in% 30:39) |> 
  dplyr::pull(total_donations) |> 
  mean() |> 
  round()
```


## Exercise 5 - Plotting and graphing

```{r}
# Using your charity dataset from the previous exercises, plot
# total_donations (y-axis) against age (x-axis)
plot(charity_data$age, charity_data$total_donations)


```

```{r}
# Using an appropriate function, plot a histogram of total donations.
# Use Help if you need to.
hist(charity_data$total_donations)


```

```{r}
# EXTENSION: If you know ggplot2, use it to create a boxplot of
# total donations grouped by gender.
library(ggplot2)
ggplot(data = charity_data, aes(x = gender, y = total_donations)) +
  geom_boxplot() +
  labs(x = "Gender", y = "Total donations") +
  theme_minimal()
 

```

```{r}
# EXTENSION: Run a pairplot to see if you can identify interesting patterns
# in the charity_data dataset.  Remember to consider the data types.
library(GGally)

charity_data <- charity_data |> 
  dplyr::mutate(
    across(c("recent_donation", "gender", "reside"), as.factor)
  )

GGally::ggpairs(charity_data)
```


## Exercise 6 - Documenting in R Markdown

```{r}
# Knit this document into a HTML file using the 'Knit' button
# View your output

```
