---
title: "1 Solutions - R Fundamentals"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this document we will briefly practice working with different vector types.  Follow the instructions in the comments of each code chunk.

## Exercise 1 - Creating and examining vectors

```{r}
# In a recent coding competition, Liz came first, Alex came second 
# and Jiena came third.

# here is a vector with their names
name <- c("Alex", "Jiena", "Liz")

# Check that this vector is a character vector of length 3


```

```{r}
# Create a corresponding numeric vector with each person's position.


```

```{r}
# Check that your vector is a numeric vector of length three.


```


## Exercise 2 - Converting vectors to factors and ordered factors 

```{r}
# Convert the 'name' character vector to a factor vector called 'name_factor'.
# Check that the conversion succeeded.


```

```{r}
# Convert your 'name' character vector to an ordered factor vector
# in increasing position order. Check that the conversion succeeded. 


```


```{r}
# EXTENSION: Use a function to display the vector in order of position (highest to lowest)?


```

## Exercise 3 - Loading and inspecting dataframes

```{r}
# Here is a URL for on online csv of data about charity donations
url <- "https://peopleanalytics-regression-book.org/data/charity_donation.csv"

# Download this into a dataframe with a name of your choice


```

```{r}
# Inspect the dataframe to find out how many rows of data there are
# and what the columns names and data types are


```

## Exercise 4 - Using functions, packages and the pipe operator

```{r}
# Load the dplyr package 


# Seek help on the functions filter() and pull() from dplyr
# What do these functions do and what are their arguments?


```

```{r}
# Using the pipe operator, write code to calculate the average amount
# donated by Rural Domestic individuals, rounded to the nearest dollar.
# Remember that == (double equals) is used for precise equality



```

```{r}
## EXTENSION: Write similar code to calculate the average amount donated 
## by males the age range of 30-39 inclusive, rounded to the nearest dollar.



```

## Exercise 5 - Plotting and graphing

```{r}
# Using your charity dataset from the previous exercises, plot
# total_donations (y-axis) against age (x-axis)


```

```{r}
# Using an appropriate function, plot a histogram of total donations.



```

```{r}
# EXTENSION: If you know ggplot2, use it to create a boxplot of
# total donations grouped by gender.



```


## Exercise 6 - Documenting in R Markdown

```{r}
# Knit this document into a HTML file using the 'Knit' button
# View your output

```
