---
title: "2A - Statistical Inference"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this document we will briefly practice some foundational concepts of statistical inference and hypothesis testing.  Follow the instructions in the comments of each code chunk.

## Exercise 1 - Mean and standard deviation of a sample

```{r}
# Download the charity donation dataset from the URL provided
url <- "https://peopleanalytics-regression-book.org/data/charity_donation.csv"



```

```{r}
# Calculate the mean total donation for this sample

# Calculate the standard deviation of the total donations for this sample


```




## Exercise 2 - Standard error and confidence interval

```{r}
# Calculate the standard error for the mean total donations of the population

# Determine the upper and lower bounds for the 95% confidence interval for the 
# mean total donations for the population


```


```{r}
# EXTENSION: Determine the upper and lower bounds for the 99% confidence interval 
# for the mean total donations for the population
# You may need to search online for an appropriate standard error multiple


```

## Exercise 3 - Loading and inspecting dataframes

```{r}
# Here is a URL for on online csv of data about charity donations
url <- "https://peopleanalytics-regression-book.org/data/charity_donation.csv"

# Download this into a dataframe with a name of your choice


```

```{r}
# Inspect the dataframe to find out how many rows of data there are
# and what the columns names and data types are


```

## Exercise 4 - Using functions, packages and the pipe operator

```{r}
# Load the dplyr package 


# Seek help on the functions filter() and pull() from dplyr
# What do these functions do and what are their arguments?


```

```{r}
# Using the pipe operator, write code to calculate the average amount
# donated by Rural Domestic individuals, rounded to the nearest dollar.
# Remember that == (double equals) is used for precise equality



```

```{r}
## EXTENSION: Write similar code to calculate the average amount donated 
## by males the age range of 30-39 inclusive, rounded to the nearest dollar.



```

## Exercise 5 - Plotting and graphing

```{r}
# Using your charity dataset from the previous exercises, plot
# total_donations (y-axis) against age (x-axis)


```

```{r}
# Using an appropriate function, plot a histogram of total donations.



```

```{r}
# EXTENSION: If you know ggplot2, use it to create a boxplot of
# total donations grouped by gender.



```


## Exercise 6 - Documenting in R Markdown

```{r}
# Knit this document into a HTML file using the 'Knit' button
# View your output

```
