---
title: "2A - Statistical Inference - SOLUTIONS"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this document we will briefly practice some foundational concepts of statistical inference and hypothesis testing.  Follow the instructions in the comments of each code chunk.

## Exercise 1 - Running a $t$-test

```{r}
# Download the charity donation dataset from the URL provided
url <- "https://peopleanalytics-regression-book.org/data/charity_donation.csv"
charity_data <- read.csv(url)

```

```{r}
# Create two vectors capturing the total donations for those who have 
# recently donated and those who have not
donations_recent <- charity_data$total_donations[charity_data$recent_donation == 1]
donations_notrecent <- charity_data$total_donations[charity_data$recent_donation == 0]
```

```{r}
# Calculate the difference in the means of these two vectors
# Round this to 2 decimal places
diff <- mean(donations_recent) - mean(donations_notrecent) 
(diff <- round(diff, 2))

```

```{r}
# Test the hypothesis that recent donors have a different mean donation amount. 
# Ensure that your test is saved as an object with a name of your choice.
(diff_test <- t.test(donations_recent, donations_notrecent)) 

```

```{r}
# EXTENSION:  Test the hypothesis that recent donors donate MORE
# (Hint: seek help on the t.test function to work out how to do this)
t.test(donations_recent, donations_notrecent, alternative = "greater")

```

## Exercise 2 - Interpreting a $t$-test

```{r}
# The results of a t-test are actually a named list.
# You can access specific elements of the list using $ (eg test$p.value)
# Return the standard error value for the difference in mean total donations. 
# Round this to 2 decimal places
(se <- diff_test$stderr |> 
   round(2))

```

```{r}
# Return the p-value and the 95% confidence interval for the population diff.
# round these to 2 decimal places 
(pval <- diff_test$p.value |> 
   round(2))

(confint <- diff_test$conf.int |> 
    round(2))
```

Use these values to write an interpretation of your $t$-test, explaining what you have observed in the sample and what this means you can infer about the population.

* The null hypothesis is that there is no difference in the mean donations.
* The observed difference in means for the sample provided was `r diff`.
* The standard error of this difference in means is `r se`.
* The 95% confidence interval for the population difference is approimately 1.96
standard errors either side of the observed value, which calculates to `r confint`.
* A difference of zero is not inside the 95% confidence interval.
* Therefore the likelihood of this difference occurring if the null hypothesis 
were true is less than 5%
* We can infer a rejection of the null hypothesis and an acceptance of the 
alternative hypothesis that there is a difference in the mean donations.
* The $p$-value of the test indicates that the likelihood of this difference 
or a greater difference occurring if the null hypothesis were true is 
approximately `r pval*100`%.


