---
title: "2A - Statistical Inference"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this document we will briefly practice some foundational concepts of 
statistical inference and hypothesis testing.  Follow the instructions 
in the comments of each code chunk.

## Exercise 1 - Running a $t$-test

```{r}
# Download the charity donation dataset from the URL provided
url <- "https://peopleanalytics-regression-book.org/data/charity_donation.csv"


```

```{r}
# Create two vectors capturing the total donations for those who have 
# recently donated and those who have not



```

```{r}
# Calculate the difference in the means of these two vectors
# Round this to 2 decimal places.



```

```{r}
# Test the hypothesis that recent donors have a different mean donation amount. 
# Ensure that your test is saved as an object with a name of your choice.



```

```{r}
# EXTENSION:  Test the hypothesis that recent donors donate MORE
# (Hint: seek help on the t.test function to work out how to do this)



```

## Exercise 2 - Interpreting a $t$-test

```{r}
# The results of a t-test are actually a named list.
# You can access specific elements of the list using $ (eg my_test$p.value)
# Return the standard error value for the difference in mean total donations. 
# Round this to 2 decimal places



```

```{r}
# Return the p-value and the 95% confidence interval for the population diff.
# round these to 2 decimal places 




```

Use these values to write an interpretation of your $t$-test, explaining what you have observed in the sample and what this means you can infer about the population.





