---
title: "2B - Linear Regression"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this document we will conduct linear regression modeling on a data set relating to graduate salaries in the United States, to try to understand economic factors which influence gradiate salaries,

## Exercise 1 - Running a simple linear regression model

```{r}
# Download the graduates dataset from the URL provided
url <- "https://peopleanalytics-regression-book.org/data/graduates.csv"


```

```{r}
# view the first few rows in the data set to acquaint yourself with the data

```

```{r}
# Determine how many observations we have in this data set

```

```{r}
# use the colnames() function to get the names of the columns in the data set

```

```{r}
# Use the cor() function to determine the correlation between unemployment rate and median salary


```

```{r}
# Consider the scale of the Unemployment_rate column.  
# Transform it into a more useful scale
# (Hint: we will want to understand the impact of a percentage point change)

```


```{r}
# Run a simple linear regression model to estimate the influence of unemployment rate
# on graduate salaries, saving your model using a name of your choice



```



## Exercise 2 - Interpreting the coefficients

```{r}
# Examine the coefficients of your saved model



```

```{r}
# Determine the 95% confidence interval for the coefficients



```

Write below an interpretation of these coefficients using the estimates and the confidence intervals which you just calculated.


```{r}
# EXTENSION:  If you are familiar with using ggplot2, create a plot of unemployment rate and
# median salary and show the estimated linear model using geom_smooth()

```

## Exercise 3 - Running a multiple linear regression model

```{r}
# Run a multiple linear regression to determine the influence of both unemployment rate
# and total graduate employees on median salary


```

```{r}
# View the coefficients and confidence intervals



```

Write below an interpretation of the coefficients of this model.


## Exercise 4 - Including categorical input variables

```{r}
# Add discipline as an input variable to the previous model you created in 
# Exercise 3.  


```


```{r}
# Inspect the results and determine which value was used as a reference for 
# discipline in the model


```


Write below your thoughts on whether any specific disciplines have a significant influence on median salary.

```{r}
# EXTENSION:  Change your reference discipline to a discipline of your choice,
# rerun the model and inspect the results.  How have they changed?

```



## Exercise 5 - Determining model fit

```{r}
# Determine the fit of your model from Exercise 4 

```

```{r}
# Determine if any of the input variables can be removed to form a
# more parsimonious model

```

```{r}
# Verify that these is no substantial difference in the fit of your
# more parsimonious model

```

```{r}
# Run a summary of your parsimonious model and comment on
# whether you are confident that it is better than a null model

```

## EXTENSION Exercise 6 - Collinearity

```{r}
# Determine if the two numerical input variables have a substantial correlation

```

```{r}
# Using the package mctest, use the imcdiag() function to test if collinearity 
# exists in your full model from Exercise 3.  Use the VIF method for this.


```
