---
title: "2B - Linear Regression Solutions"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this document we will conduct linear regression modeling on a data set relating to graduate salaries in the United States, to try to understand economic factors which influence graduate salaries.

## Exercise 1 - Running a simple linear regression model

```{r}
# Download the graduates dataset from the URL provided
url <- "https://peopleanalytics-regression-book.org/data/graduates.csv"
graduates <- read.csv(url)

```

```{r}
# view the first few rows in the data set to acquaint yourself with the data
head(graduates)
```


```{r}
# Determine how many observations we have in this data set
nrow(graduates)
```

```{r}
# use the colnames() function to get the names of the columns in the data set
colnames(graduates)
```

```{r}
# Use the cor() function to determine the correlation between unemployment rate and median salary
cor(graduates$Unemployment_rate, graduates$Median_salary)

```

```{r}
# Consider the scale of the Unemployment_rate column.  
# Transform it into a more useful scale
# (Hint: we will want to understand the impact of a percentage point change)
graduates$Unemployment_rate <- 100*graduates$Unemployment_rate
```


```{r}
# Run a simple linear regression model to estimate the influence of unemployment rate
# on graduate salaries, saving your model using a name of your choice
model <- lm(data = graduates, formula = Median_salary ~ Unemployment_rate)


```



## Exercise 2 - Interpreting the coefficients

```{r}
# Examine the coefficients of your saved model
model$coefficients


```

```{r}
# Determine the 95% confidence interval for the coefficients
confint(model)


```

Write below an interpretation of these coefficients using the estimates and the confidence intervals which you just calculated.

* Unemployment rate has a statistically significant influence on median graduate salary
* Full employment is associated with a median salary of between \$63425 and \$76769 on average
* An additional percentage point in unemployment rate is associated with a reduction in median salary of between \$1212 and \$3419 on average

```{r}
# EXTENSION:  If you are familiar with using ggplot2, create a plot of unemployment rate and
# median salary and show the estimated linear model using geom_smooth()
library(ggplot2)

ggplot(data = graduates, aes(x = Unemployment_rate, y = Median_salary)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Unemployment rate (%)",
       y = "Median salary ($)") +
  theme_minimal()
```

## Exercise 3 - Running a multiple linear regression model

```{r}
# Run a multiple linear regression to determine the influence of both unemployment rate
# and total graduate employees on median salary
model2 <- lm(Median_salary ~ Unemployment_rate + Total, graduates)

```

```{r}
# View the coefficients and confidence intervals
model2$coefficients

confint(model2)


```

Write below an interpretation of the coefficients of this model.


## Exercise 4 - Including categorical input variables

```{r}
# Add discipline as an input variable to the previous model you created in 
# Exercise 3.  
model3 <- lm(Median_salary ~ Unemployment_rate + Total + Discipline, 
             graduates)

```


```{r}
# Inspect the results and determine which value was used as a reference for 
# discipline in the model
confint(model3)

# view all unique values for discipline
unique(graduates$Discipline)

# it appears that the first discipline in alphabetical order has been used
# as reference (Agriculture & Natural Resources)

```

Write below your thoughts on whether any specific disciplines have a significant influence on median salary.

Compared to a reference discipline of Agriculture & Natural Resources, the following disciplines have a significant positive influence on median salary (assuming similar unemployment rates and total graduate numbers):

* Computers & Mathematics (Yay for R programmers!) 
* Engineering
* Physical Sciences

and the following have a significant negative influence:

* Education

```{r}
# EXTENSION:  Change your reference discipline to a discipline of your choice,
# rerun the model and inspect the results.  How have they changed?
graduates$Discipline <- as.factor(graduates$Discipline) |> 
  relevel(ref = "Computers & Mathematics")

model3 <- lm(Median_salary ~ Unemployment_rate + Total + Discipline, 
             graduates)

confint(model3)
```

Because we have referenced on one of the higher salary disciplines, we now see a lot more disciplines which have a  significantly negative influence on median salary.


## Exercise 5 - Determining model fit

```{r}
# Determine the fit of your model from Exercise 4 
summary(model3)$r.squared
```

```{r}
# Determine if any of the input variables can be removed to form a
# more parsimonious model

# we know from Exercise 4 that we can remove Total 
model4 <- lm(Median_salary ~ Unemployment_rate + Discipline, graduates)
```

```{r}
# Verify that these is no substantial difference in the fit of your
# more parsimonious model
summary(model4)$r.squared
```

```{r}
# Run a summary of your parsimonious model 
summary(model4)

```

The $p$-value of the F-statistic is very small, indicating high confidence that the model is better than a null model.

## EXTENSION Exercise 6 - Model diagnostics

```{r}
# Determine if the two numerical input variables have a significant correlation
cor.test(graduates$Unemployment_rate, graduates$Total)
```

```{r}
# Using the package mctest, use the imcdiag() function to test if collinearity 
# exists in your full model from Exercise 3.  Use the VIF method for this.
library(mctest)

mctest::imcdiag(model3, method = "VIF")

```


```{r}
# View a density plot of the Median salaries in the data set - what do you observe?
density(graduates$Median_salary) |> 
  plot(main = "Median Salary Density")
```

The distribution appears to be somewhat left-skewed due to a tail of high median salaries.

```{r}
# use the qqnorm() function to determine how the residuals of your model 
# compare to a normal distribution
# comment on your observations
# consider re-running your model to address any concerns you have

qqnorm(model4$residuals)

```

Observations:
* Linear regression assumes a normal distribution of residuals
* Therefore we want this to be as close to a straight line as possible
* Two high median salary outliers appear to be affecting this
* We could consider removing these outliers to help build a more reliable model



```{r}
# remove highest two median salaries

library(dplyr)
adjusted_graduates <- graduates |> 
  dplyr::arrange(Median_salary) |> 
  head(nrow(graduates) - 2)

# rerun model
model5 <- lm(Median_salary ~ Unemployment_rate + Discipline, 
             adjusted_graduates)

# test residuals now
qqnorm(model5$residuals)
```
Healthier looking residual distribution.

```{r}
# view results
summary(model5)
```

Overall fit has improved, with similar significant variables, but some changes to coefficients.