---
title: "03A - Binomial Logistic Regression Solutions"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this document we will conduct binomial logistic regression modeling on a data set relating to recruiting data and whether a person is hired or not hired based on test scores, interviews, and gender.

## Exercise 1 - Running a simple binomial logistic regression model

```{r}
# Download the recruiting dataset from the URL provided
url <- "https://peopleanalytics-regression-book.org/data/recruiting.csv"
recruiting <- read.csv(url)

```

```{r}
# view the first few rows in the data set to acquaint yourself with the data
head(recruiting)
```

```{r}
# What do you think the datatypes are?
# gender:
# sat: 
# apttest:
# int1:
# int2:
# int3:
# hired:
```


```{r}
# Perform any necessary datatype adjustments
recruiting$gender <- as.factor(recruiting$gender)
recruiting$hired <- as.factor(recruiting$hired)
recruiting$int1 <- ordered(recruiting$int1,
                           levels = 1:5)
recruiting$int2 <- ordered(recruiting$int2,
                           levels = 1:5)
recruiting$int3 <- ordered(recruiting$int3,
                           levels = 1:5)
```


```{r}
# Run a simple binomial logistic regression model to estimate the influence of SAT scores
# on hire success, saving your model using a name of your choice
model <- glm(data = recruiting, formula = hired ~  sat, family = "binomial")

```



## Exercise 2 - Interpreting the coefficients

```{r}
# Examine the coefficients of your saved model
summary(model)$coefficients

```

```{r}
# Take a look at the odds ratio for our independent variable. How do you interpret this?
exp(model$coefficients)

```

```{r}
# Do you think this is a good model?
DescTools::PseudoR2(model)
```

## Exercise 3 - Running a multiple binomial logistic regression model

```{r}
# What variables look like they might be interesting to include?
GGally::ggpairs(recruiting)

```

```{r}
# let's try modeling hired based on apttest and gender

model2 <- glm(hired ~ apttest + gender, data = recruiting, family = "binomial")

```

```{r}
# View the coefficients and log odds
summary(model2)$coefficients
exp(model2$coefficients)
```

```{r}
# Assess model fit - is this better than the simple model?
DescTools::PseudoR2(model2)
AIC(model)
AIC(model2)
```

