---
title: "03A - Binomial Logistic Regression Solutions"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this document we will conduct binomial logistic regression modeling on a data set relating to recruiting data and whether a person is hired or not hired based on test scores, interviews, and gender.  Data fields: 

  - gender: gender
  - sat: SAT test score of applicant
  - gpa: GPA of applicant
  - apttest: Score on an aptitude test given to applicant
  - int1: applicant rating given by one of two line manager interviewers (1-5 where high is positive)
  - int2: applicant rating given by second of two line manager interviewers (1-5 where high is positive)
  - int3: applicant rating given by HR interviewer (1-5 where high is positive)
  - hired: hired (1) or not hired (0)

## Exercise 1 - Running a simple binomial logistic regression model

```{r}
# Download the recruiting dataset from the URL provided
url <- "https://peopleanalytics-regression-book.org/data/recruiting.csv"
recruiting <- read.csv(url)

```

```{r}
# view the first few rows in the data set to acquaint yourself with the data
head(recruiting)
```

```{r}
# What do you think the datatypes are?
# gender: binary factor
# sat: numeric
# gpa: numeric
# apttest: numeric
# int1: ordered factor
# int2: ordered factor
# int3: ordered factor
# hired: binary factor
```


```{r}
# Perform any necessary datatype adjustments
recruiting$gender <- as.factor(recruiting$gender)
recruiting$hired <- as.factor(recruiting$hired)
recruiting$int1 <- ordered(recruiting$int1,
                           levels = 1:5)
recruiting$int2 <- ordered(recruiting$int2,
                           levels = 1:5)
recruiting$int3 <- ordered(recruiting$int3,
                           levels = 1:5)
```

```{r}
# get to the know the data a bit.  what is the split of hired/not hired,
# do you notice differences between the interviewers, etc
summary(recruiting)
GGally::ggpairs(recruiting)
```

```{r}
# Run a simple binomial logistic regression model to estimate the influence of SAT scores
# on hire success, saving your model using a name of your choice
model <- glm(data = recruiting, formula = hired ~  sat, family = "binomial")

```


## Exercise 2 - Interpreting the coefficients

```{r}
# Examine the coefficients of your saved model
summary(model)$coefficients

```

```{r}
# Take a look at the odds ratio for our independent variable. How do you interpret this?
exp(model$coefficients)

```

```{r}
# Do you think this is a good model?
DescTools::PseudoR2(model)
```

## Exercise 3 - Running a multiple binomial logistic regression model

```{r}
# What variables look like they might be interesting to include?
GGally::ggpairs(recruiting)

```

```{r}
# make a new model with gender and apttest
model2 <- glm(hired ~ apttest + gender,  data = recruiting, family = "binomial")
```

```{r}
# View the coefficients and log odds.
summary(model2)$coefficients
exp(model2$coefficients)
```


```{r}
# Assess model fit - is this better than the simple model?
DescTools::PseudoR2(model2)
AIC(model)
AIC(model2)
```

Write a few sentences on the data and the results of your model.  Odds can be difficult to explain - can you try to 
explain to a stakeholder what this means for a candidate with a 60 on their aptitude test vs a candidate with a 70?

Each additional point in the aptitude tests results in a ~6.5% increase in the odds of being hired.
The low Pseudo R^2 value means that we are only explaining a small amount of what leads to a person being hired.

A candidate with a 70 has 10 more points than a candidate with 60.  Each point relates to a 6.5% increase in odds, so all
else being equal, the person with a 70 has 1.065^10 = ~88% increased odds of being hired vs the candidate with a 60.
