---
title: "03A - Binomial Logistic Regression"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this document we will conduct binomial logistic regression modeling on a data set relating to recruiting data and whether a person is hired or not hired based on test scores, interviews, and gender.  Data fields: 

  - gender: gender
  - sat: SAT test score of applicant
  - gpa: GPA of applicant
  - apttest: Score on an aptitude test given to applicant
  - int1: applicant rating given by one of two line manager interviewers (1-5 where high is positive)
  - int2: applicant rating given by second of two line manager interviewers (1-5 where high is positive)
  - int3: applicant rating given by HR interviewer (1-5 where high is positive)
  - hired: hired (1) or not hired (0)

## Exercise 1 - Running a simple binomial logistic regression model

```{r}
# Download the recruiting dataset from the URL provided
url <- "https://peopleanalytics-regression-book.org/data/recruiting.csv"
recruiting <- read.csv(url)

```

```{r}
# view the first few rows in the data set to acquaint yourself with the data

```

```{r}
# What do you think the datatypes are?
# gender: 
# sat: 
# gpa: 
# apttest: 
# int1: 
# int2:
# int3: 
# hired: 
```


```{r}
# Perform any necessary datatype adjustments

```

```{r}
# get to the know the data a bit.  what is the split of hired/not hired,
# do you notice differences between the interviewers, etc
# Hint: summary(), GGally::ggpairs() are good first steps

```

```{r}
# Run a simple binomial logistic regression model to estimate the influence of SAT scores
# on hire success, saving your model using a name of your choice


```


## Exercise 2 - Interpreting the coefficients

```{r}
# Examine the coefficients of your saved model


```

```{r}
# Take a look at the odds ratio for our independent variable. How do you interpret this?


```

```{r}
# Do you think this is a good model?

```

## Exercise 3 - Running a multiple binomial logistic regression model

```{r}
# What variables look like they might be interesting to include?


```

```{r}
# make a new model with gender and apttest

```

```{r}
# View the coefficients and log odds.

```


```{r}
# Assess model fit - is this better than the simple model?

```

Write a few sentences on the data and the results of your model.  Odds can be difficult to explain - can you try to 
explain to a stakeholder what this means for a candidate with a 60 on their aptitude test vs a candidate with a 70?
