---
title: "03B-Ordinal_regression"
output: html_document
---

  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this document we will conduct ordinal logistic regression on fictional data on employee performance evaluation metrics for a group of salespeople. 

  - sales: annual sales of the individual in millions of dollars
  - new_customers: number of new customers acquired by individual
  - region: region individual works in (North, South, East, West)
  - gender: gender of individual
  - rating: performance rating of individual (1 = Low, 2 = Middle, 3 = High)

## Exercise 1 - Running a simple ordinal logistic regression model

```{r}
# Download the employee performance dataset from the URL provided
url <- "https://peopleanalytics-regression-book.org/data/employee_performance.csv"
perf <- read.csv(url)

```

```{r}
# view the first few rows in the data set to acquaint yourself with the data

```

```{r}
# What do you think the datatypes are?
# sales:
# new_customers:
# region:
# gender: 
# rating: 
```


```{r}
# Perform any necessary datatype adjustments

```

```{r}
# Take a look at the pairs plot. What do you notice?
# Hint: Use GGally::ggpairs()

```

```{r}
# Run a simple ordinal logistic regression model to understand sales influence
# on rating, saving your model using a name of your choice

```



## Exercise 2 - Interpreting the coefficients

```{r}
# Examine the coefficients of your saved model


```

```{r}
# Add p-values and odds ratios, and view
# What do you notice?



```

Write a sentence on the model results above.


```{r}
# Do you think this is a good model?  Use the lipsitz test.

```
Write a sentence on the results of this test.


## Exercise 3 - Add more information to the model

```{r}
# Let's try a model to see how sales, new_customers, and gender impact rating

```

```{r}
# Add p-values and odds ratios, and view
# What do you notice?

```

Write a few sentences on the results of the model.


```{r}
# Do you think this is a good model?

```

Write a sentence on the results of the Lipsitz test.


# Exercise 4: Test the proportional odds assumption

```{r}
# First try the method where we compare the binary models to each other and use
# our judgement

```

Using your best judgement, do you think the proportional odds assumption is met?



```{r}
# Alternatively with the Brant-Wald test


```

Write your interpretation of the brant-wald test results.

