---
title: "5 - Creating and Visualizing Graphs"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this document we will practice creating and examining graph objects in R, and visualizing those graphs using common visualization packages.

## Exercise 1 - Creating an undirected graph object

```{r}
# Download the karate dataset from the URL provided
url <- "https://ona-book.org/data/karate.csv"

karate_edgelist <- read.csv(url)
```

```{r}
# this dataset is the edgelist from a study of a karate club by Wayne W. Zachary in 1977
# an edge between two vertices means that the two members socialized outside the club

# create an undirected graph from the edgelist
library(igraph)

karate_graph <- igraph::graph_from_data_frame(
  karate_edgelist,
  directed = FALSE
)
```

```{r}
# View the graph object - how many vertices and edges are there?
karate_graph
```

```{r}
# View the vertex set
V(karate_graph)
```

```{r}
# View the edge set
E(karate_graph)

```


## Exercise 2 - Creating a directed graph object

```{r}
# Download the schoolfriends edgelist from the URL provided
url <- "https://ona-book.org/data/schoolfriends_edgelist.csv"

schoolfriends_edgelist <- read.csv(url)

```

```{r}
# This network is from a research project on French high school children
# This edgelist gives two types of relationships.
# The first type is a reported friendship, where the first person reported
# the second person as a a friend in a survey.
# The second type is a known Facebook friendship connection between the two.

# filter the edgelist to just the reported friendships.
library(dplyr)

schoolfriends_rp_edgelist <- schoolfriends_edgelist |> 
  dplyr::filter(type == "reported")
```

```{r}
# create an appropriate graph object from this filtered edgelist. 
schoolfriends_rp_graph <- igraph::graph_from_data_frame(
  schoolfriends_rp_edgelist,
  directed = TRUE
)


```


```{r}
# Examine the graph object - how many vertices and edges does it have
schoolfriends_rp_graph
```




## Exercise 3 - Comparing graphs

```{r}
# Filter the schoolfriends data set to just the facebook friendships
schoolfriends_fb_edgelist <- schoolfriends_edgelist |> 
  dplyr::filter(type == "facebook")

```

```{r}
# Create and examine an appropriate graph object from this filtered data set
(schoolfriends_fb_graph <- igraph::graph_from_data_frame(
  schoolfriends_fb_edgelist,
  directed = FALSE
))


```

Write below a comparison of the basic properties of the two types of `schoolfriends` networks:

1. The reported friendship graph is directed, whereas the Facebook friendship graph is undirected.
2. The reported friendship graph has 134 vertices, compared to 156 vertices for the Facebook friendship graph, indicting that it is more more students had Facebook friendships versus reported friendships.
3. There are 668 edges in the reported friendship graph, but 1437 in the Facebook friendship graph.  This indicates that Facebook friendships are more common than reported friendships.


## Exercise 4 - Loading a graph with edge properties

```{r}
# Load the workfrance_edgelist data set from the URL provided.
# This is an edgelist of a network of interactions between office employees
# in various departments based on how much time they spent together during a given day
url <- "https://ona-book.org/data/workfrance_edgelist.csv"
workfrance_edgelist <- read.csv(url)

```

```{r}
# Create a undirected graph object from this edgelist including any edge properties
workfrance_graph <- igraph::graph_from_data_frame(
  workfrance_edgelist,
  directed = FALSE
)


```

```{r}
# verify that there is an edge property called mins and show the first ten values
# of this edge property
E(workfrance_graph)$mins[1:10]
```

## Exercise 5 - Loading a graph with vertex properties

```{r}
# Load the workfrance_vertices data set from the URL provided
url <- "https://ona-book.org/data/workfrance_vertices.csv"
workfrance_vertices <- read.csv(url)

```

```{r}
# Recreate the undirected graph object with both edge and vertex properties
workfrance_graph <- igraph::graph_from_data_frame(
  workfrance_edgelist,
  vertices = workfrance_vertices,
  directed = FALSE
)


```

```{r}
# find the vertex properties in your new graph
# choose one vertex property and show the first ten values
V(workfrance_graph)$dept[1:10]

```


## Exercise 6 - Examining graph properties

```{r}
# Identify the data types of each of the edge and vertex properties in your
# workfrance graph
workfrance_graph
```

```{r}
# Use the graph object to find the mean number of mins for all edges in your 
# workfrance graph
mean(E(workfrance_graph)$mins)


```

```{r}
# EXTENSION - Create a subgraph of all vertices in the 'SSI' department
# calculate the mean number of mins for all edges in this subgraph
# Hint: try the igraph::induced_subgraph() function

SSI_vertices <- V(workfrance_graph)[V(workfrance_graph)$dept == 'SSI']

SSI_subgraph <- igraph::induced_subgraph(
  workfrance_graph,
  vids = SSI_vertices
)

mean(E(SSI_subgraph)$mins)
```


## Exercise 7 - Plotting a network graph

```{r}
# Run a basic plot of the undirected karate graph
plot(karate_graph)

```

```{r}
# Remove all labels except 'Mr Hi' and 'John A'
V(karate_graph)$label <- ifelse(V(karate_graph)$name %in% c('Mr Hi', 'John A'),
                                V(karate_graph)$name,
                                "")

plot(karate_graph)
```


```{r}
# Re-do the layout using the Kamada-Kawaii layout
# Set a seed of 123 to control randomness
# Check whether your neighbor generated an identical plot
set.seed(123)
karate_graph$layout <- igraph::layout_with_kk(karate_graph)
plot(karate_graph)
```

## Exercise 8 - Plotting a network graph (EXTENSION)

```{r}
# Plot the undirected workfrance network graph without labels and using a 
# Fruchterman-Reingold force directed layout.  Use a seed of 123.
set.seed(123)
workfrance_graph$layout <- igraph::layout_with_fr(workfrance_graph)
V(workfrance_graph)$label <- ""
plot(workfrance_graph)

```

```{r}
# Resize the vertices to size 5
V(workfrance_graph)$size <- 5
plot(workfrance_graph)
```
```{r}
# Resize the vertices to size 5
V(workfrance_graph)$size <- 5
plot(workfrance_graph)
```


```{r}
# Try to color the vertices of the graph by the dept of the vertex
V(workfrance_graph)$color <- factor(V(workfrance_graph)$dept)
plot(workfrance_graph)

```


## Exercise 9 - Plotting a network graph using `ggraph` (EXTENSION)

```{r}
# Create a basic Plot of the workfrance undirected network using ggraph
# use a Kamada-Kawaii layout and set a seed of 123
library(ggraph)

set.seed(123)
ggraph(workfrance_graph, layout = "kk") +
  geom_edge_link(color = "grey") +
  geom_node_point(size = 3) +
  theme_void()

```

```{r}
# Redo the plot again using an aesthetic to color the vertices by dept
set.seed(123)
ggraph(workfrance_graph, layout = "kk") +
  geom_edge_link(color = "grey") +
  geom_node_point(size = 3, aes(color = dept)) +
  theme_void()
```

```{r}
# (EXTENSION) Try to scale the edge by the mins edge property
set.seed(123)
ggraph(workfrance_graph, layout = "kk") +
  geom_edge_link(aes(color = mins)) +
  geom_node_point(size = 3, aes(color = dept),
                  show.legend = FALSE) +
  theme_void()

```


## Exercise 10 - Interactive network visualization using `networkD3` (EXTENSION)

```{r}
# Create a group in the karate graph containing Mr Hi and John A
V(karate_graph)$group <- ifelse(V(karate_graph)$name %in% c("Mr Hi", "John A"), 1, 0)
```

```{r}
# Convert the karate graph for the purposes of using networkD3
# Use you new group in the conversion
library(networkD3)

karate_d3 <- networkD3::igraph_to_networkD3(karate_graph, 
                                            group = V(karate_graph)$group)
```

```{r}
# Run a force network visualization of the karate graph in networkD3 
networkD3::forceNetwork(Links = karate_d3$links, 
                        Nodes = karate_d3$nodes, 
                        NodeID = "name",   
                        Source = "source",
                        Target = "target", 
                        Group = "group",
                        opacity = 1)

```