---
title: "5 - Creating and Visualizing Graphs"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this document we will practice creating and examining graph objects in R, and visualizing those graphs using common visualization packages.

## Exercise 1 - Creating an undirected graph object

```{r}
# Download the karate dataset from the URL provided
url <- "https://ona-book.org/data/karate.csv"


```

```{r}
# this dataset is the edgelist from a study of a karate club by Wayne W. Zachary in 1977
# an edge between two vertices means that the two members socialized outside the club

# create an undirected graph from the edgelist

```

```{r}
# View the graph object - how many vertices and edges are there?

```

```{r}
# View the vertex set

```

```{r}
# View the edge set


```


## Exercise 2 - Creating a directed graph object

```{r}
# Download the schoolfriends edgelist from the URL provided
url <- "https://ona-book.org/data/schoolfriends_edgelist.csv"



```

```{r}
# This network is from a research project on French high school children
# This edgelist gives two types of relationships.
# The first type is a reported friendship, where the first person reported
# the second person as a a friend in a survey.
# The second type is a known Facebook friendship connection between the two.

# filter the edgelist to just the reported friendships.

```

```{r}
# create an appropriate graph object from this filtered edgelist. 



```


```{r}
# Examine the graph object - how many vertices and edges does it have

```


## Exercise 3 - Comparing graphs

```{r}
# Filter the schoolfriends data set to just the facebook friendships


```

```{r}
# Create and examine an appropriate graph object from this filtered data set



```

Write below a comparison of the basic properties of the two types of `schoolfriends` networks:



## Exercise 4 - Loading a graph with edge properties

```{r}
# Load the workfrance_edgelist data set from the URL provided.
# This is an edgelist of a network of interactions between office employees
# in various departments based on how much time they spent together during a given day
url <- "https://ona-book.org/data/workfrance_edgelist.csv"


```

```{r}
# Create a undirected graph object from this edgelist including any edge properties



```

```{r}
# verify that there is an edge property called mins and show the first ten values
# of this edge property

```

## Exercise 5 - Loading a graph with vertex properties

```{r}
# Load the workfrance_vertices data set from the URL provided
url <- "https://ona-book.org/data/workfrance_vertices.csv"


```

```{r}
# Recreate the undirected graph object with both edge and vertex properties



```

```{r}
# find the vertex properties in your new graph
# choose one vertex property and show the first ten values


```


## Exercise 6 - Examining graph properties

```{r}
# Identify the data types of each of the edge and vertex properties in your
# workfrance graph



```

```{r}
# Use the graph object to find the mean number of mins for all edges in your 
# workfrance graph



```

```{r}
# EXTENSION - Create a subgraph of all vertices in the 'SSI' department
# calculate the mean number of mins for all edges in this subgraph
# Hint: try the igraph::induced_subgraph() function


```


## Exercise 7 - Plotting a network graph using `igraph`

```{r}
# Run a basic plot of the undirected karate graph



```

```{r}
# Remove all labels except 'Mr Hi' and 'John A'



```


```{r}
# Re-do the layout using the Kamada-Kawaii layout
# Set a seed of 123 to control randomness
# Check whether your neighbor generated an identical plot

```

## Exercise 8 - Plotting a network graph using `igraph` (EXTENSION)

```{r}
# Plot the undirected workfrance network graph without labels and using a 
# Fruchterman-Reingold force directed layout.  Use a seed of 123.


```

```{r}
# Resize the vertices to size 5

```

```{r}
# Try to color the vertices of the graph by the dept of the vertex


```


## Exercise 9 - Plotting a network graph using `ggraph` (EXTENSION)

```{r}
# Create a basic Plot of the workfrance undirected network using ggraph
# use a Kamada-Kawaii layout and set a seed of 123


```

```{r}
# Redo the plot again using an aesthetic to color the vertices by dept

```

```{r}
# (EXTENSION) Try to scale the edge by the mins edge property


```


## Exercise 10 - Interactive network visualization using `networkD3` (EXTENSION)

```{r}
# Create a group in the karate graph containing Mr Hi and John A


```

```{r}
# Convert the karate graph for the purposes of using networkD3
# Use you new group in the conversion

```

```{r}
# Run a force network visualization of the karate graph in networkD3 


```
