---
title: "Network Analysis II"
author: "<b>rstudio::</b>conf(2022)"
output:
  xaringan::moon_reader:
    css: "style.css"
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
options(knitr.kable.NA = '')
library(ggraph)
library(igraph)
```

class: left, middle, rstudio-logo, bigfont

## Aim of this module

&#9989; Review basic network metrics
  - Learn about paths and distance
  - Review network summary metrics
  
&#9989; Understand relationships between nodes
  - Vertex importance and centrality
  - Community detection methods
  - Assortativity and Similarity
  
  
---
class: left, middle, rstudio-logo, bigfont

## Paths

Consider the below graphs.  How can we get from A to C?

```{r, echo = FALSE, fig.align = "center", out.height = 400}
g <- data.frame(
  from = c("A","A","A","B","B","C","E","B"),
  to = c("B","D","E","C","D","D","F","A")
)
d1 <- igraph::graph_from_data_frame(g, directed = TRUE)

d2 <- igraph::graph_from_data_frame(g, directed = FALSE)

p1 <-
  ggraph::ggraph(d1, layout = "kk") +
  geom_edge_link(color = "black", arrow = arrow()) +
  geom_node_label(fill = "lightblue", aes(label = name)) +
  ggtitle("Directed") +
  theme_void()

p2 <-
  ggraph::ggraph(d2, layout = "kk") +
  geom_edge_link(color = "black", alpha = 0.7) +
  geom_node_label(fill = "lightblue", aes(label = name)) +
  ggtitle("Undirected") +
  theme_void()

gridExtra::grid.arrange(p1, p2, nrow = 1)
```

---
class: left, middle, rstudio-logo

## Paths, continued

There is only 1 way in the directed graph, but many in the undirected graph.

```{r, echo = FALSE, fig.align = "right", out.width = "50%"}
library(kableExtra)

d <- data.frame(
  Directed = c(
    "A->B, B->C",
    " ",
    " ",
    " "
  ),
  Undirected = c(
    "A->B, B->C",
    "A->D, D->C",
    "A->B, B->D, D->C",
    "A->D, D->B, B->C"
  )
)

kbl(d) |>
  kable_minimal()
```


---
class: left, middle, rstudio-logo

## Distance

The distance between two nodes is the sum of the weights of the edges traversed in the path. For graphs without a weight property, every edge is assumed to have weight 1.

```{r, echo = FALSE, out.height = "300", out.width = "400", fig.align = "center", out.extra='style="float:left; padding-right:40px"'}
g <- data.frame(
  from = c("A","A","A","B","B","C","E"),
  to = c("B","D","E","C","D","D","F"),
  weight = c(3,4,2,1,6,5,4)
)

d1 <- igraph::graph_from_data_frame(g, directed = FALSE)

ggraph::ggraph(d1) +
  geom_edge_link(color = "grey", aes(label = weight)) +
  geom_node_label(fill = "lightblue", aes(label = name)) +
  theme_void()
```

```{r, echo = TRUE, fig.align = "center", out.extra='style="float:right"'}
distances(d1,weights = NULL)
```


---
class: left, middle, rstudio-logo

## Network size

How large is our network?

```{r, echo = FALSE, out.height = "400", out.width = "400", fig.align = "center", out.extra='style="float:left; padding-right:40px"'}

sge <- data.frame(
  from = c("A","B","B","C","E"),
  to = c("B","C","D","D","F")
)

sg <- igraph::graph_from_data_frame(sge, directed = FALSE)

ggraph::ggraph(sge, layout = "kk") +
  geom_edge_link(color = "grey") +
  geom_node_label(fill = "lightblue", aes(label = name)) +
  theme_void()
```

```{r, echo = TRUE, fig.align = "center", out.extra='style="float:right"'}
# Number of nodes
vcount(sg)

# Number of edges
ecount(sg)
```


---
class: left, middle, rstudio-logo

## Network density

Network density is defined as the number of actual edges divided by the number of possible edges. A graph with density of 1 is a *complete* graph, and a graph with low density is considered *sparse*.


```{r, echo = FALSE, out.height = "400", out.width = "400", fig.align = "center", out.extra='style="float:left; padding-right:40px"'}

ggraph::ggraph(sge, layout = "kk") +
  geom_edge_link(color = "grey") +
  geom_node_label(fill = "lightblue", aes(label = name)) +
  theme_void()
```

  -  Actual Ties: 5
  -  Possible Ties = $\frac{N(N-1)}{2} = 15$
```{r, echo = TRUE, fig.align = "center", out.extra='style="float:right"'}
# Network density
edge_density(sg)
```

---
class: left, middle, rstudio-logo

## Components

It can be useful to know the number of **components** in a network.  A component is defined as the number of disconnected groups in a network.

```{r, echo = FALSE, out.height = "400", out.width = "400", fig.align = "center", out.extra='style="float:left; padding-right:40px"'}

ggraph::ggraph(sge, layout = "kk") +
  geom_edge_link(color = "grey") +
  geom_node_label(fill = "lightblue", aes(label = name)) +
  theme_void()
```

```{r}
# Number of components
components(sg)
```

---
class: left, middle, rstudio-logo

## Network diameter

Network diameter is defined as the longest of the shortest paths in a network.  This is mostly useful in a *connected graph* where all nodes can connect in some way.  We have an *unconnected graph*, so this metric isn't particularly useful.

```{r, echo = FALSE, out.height = "300", out.width = "400", fig.align = "center", out.extra='style="float:left; padding-right:40px"'}

ggraph::ggraph(sge, layout = "kk") +
  geom_edge_link(color = "grey") +
  geom_node_label(fill = "lightblue", aes(label = name)) +
  theme_void()
```

```{r}
diameter(sg)
```

---
class: left, middle, rstudio-logo, bigfont

## Vertex importance and centrality

In networks involving people, understanding vertex imporance (a person's importance) is of significant interest.  Questions you might consider:

  - Who serves as a "bridge" between various parts of an organization?  If these people depart, you can end up with "islands" of people who feel disconnected.
  - Who has the highest number of connections?  This might be a good person to consider for a leadership role, assuming they are viewed positively by their connections.
  - Who is connected to a lot of well-connected people?  These people can be helpful for facilitating introductions.

---
class: left, middle, rstudio-logo, bigfont

## Types of centrality (1/2)

**Degree centrality** is simply the count of connections for each node.

**Closeness centrality** is a measure of how central or close a node is to other nodes.  Information spreads quickly through a network if it starts with those with high closeness centrality.  To calculate closeness centrality for a node $v$:

1.  Take all nodes connected to $v$ and calculate the distance between each node and $v$. 
2.  Take the average and invert it (so that higher means closer)

---
class: left, middle, rstudio-logo, bigfont
## Types of centrality (2/2)

**Betweenness centrality** is a measure of how important a node is in the overall connectedness of the network.  To derive the betweenness centrality of a node $v$:

1.  Take any pairs of nodes that are not $v$ and calculate the number of shortest paths between them.
2.  Determine how many of those paths pass through $v$.
3.  Divide 2 by 1 and sum across all pairs of nodes in the network.
  
  
---
class: left, middle, rstudio-logo

## Back to our undirected graph (1/2)

Let's create the graph together and learn how to work with it in R. We'll use the `tidygraph` package for a more convenient way to work with `igraph`.

```{r}
# our edgelist
edges_gr <- data.frame(
  from = c("A","A","A","B","B","C","E","B"),
  to = c("B","D","E","C","D","D","F","A")
)

# create a graph from the datafram using igraph
gr <- igraph::graph_from_data_frame(edges_gr, directed = FALSE)

# the tidygraph package gives us a really simple way to work with 
gr_tidy <- igraph::simplify(gr) |> # remove the A->B/B->A duplication
  tidygraph::as_tbl_graph()

```

---
class: left, middle, rstudio-logo

## Back to our undirected graph (1/2)

```{r}
gr_tidy
```
---
class: left, middle, rstudio-logo

## Centrality of our undirected graph (2/2)

Now we can use convenient `tidyverse` functions and create a quick table of centrality values for each node.

```{r, echo = TRUE, fig.align = "center"}
c_vals <-
  gr_tidy |>
  dplyr::rename(NODE = name) |>
  dplyr::mutate(
    DEGREE_CENT = tidygraph::centrality_degree(),
    BTWN_CENT = tidygraph::centrality_betweenness(),
    CLOSE_CENT = tidygraph::centrality_closeness()
  ) |>
  kbl() |>
  kable_minimal()
```

---
class: left, middle, rstudio-logo

## Centrality of our undirected graph (2/2)

Now we can use convenient `tidyverse` functions and create a quick table of centrality values for each node.

```{r, echo = FALSE, fig.align = "center"}
c_vals
```

---
class: left, middle, rstudio-logo, bigfont

## Cliques (1/2)

Cliques are subsets of vertices in an undirected graph whose induced subgraph is complete (has an edge density of 1).

How many cliques are in our network?

```{r, echo = FALSE, out.height = "300", out.width = "400", fig.align = "center"}
ggraph::ggraph(d1) +
  geom_edge_link(color = "grey") +
  geom_node_label(fill = "lightblue", aes(label = name)) +
  theme_void()

```

---
class: left, middle, rstudio-logo

## Cliques (2/2)

```{r, echo = FALSE, out.height = "400", out.width = "400", fig.align = "center", out.extra='style="float:left; padding-right:40px"'}

ggraph::ggraph(d1) +
  geom_edge_link(color = "grey") +
  geom_node_label(fill = "lightblue", aes(label = name)) +
  theme_void()
```

```{r}
# Size of the largest clique
clique_num(d1)

# Number of cliques between size 2 and 3
length(cliques(d1, min = 2, max = 3))

```

---
class: left, middle, rstudio-logo

## Community detection

Network communities are "hidden" subsets that exist within a network.  Through community detection methods, we can sometimes uncover interesting similarities between nodes that we wouldn't have otherwise noticed.  

---
class: left, middle, rstudio-logo

## Louvain algorithm (1/2)

One commonly used community detection algorithm is the Louvain algorithm, which partitions the graph into subsets of vertices by trying to maximize the *modularity* of the graph. Modularity measures how dense the connections are within subsets of vertices in a graph by comparing the density to that which would be expected by a random graph.  Modularity ranges from -0.5 to 1 and any positive value indicates the vertices inside the subgroups are more densely connected than would be expected by chance.

```{r, echo = FALSE}

url <- "https://ona-book.org/data/schoolfriends_edgelist.csv"
schoolfriends <- read.csv(url)

sf_gr <- igraph::graph_from_data_frame(schoolfriends, directed = FALSE)
```

```{r}
# find Louvain communities
communities <- cluster_louvain(sf_gr)

# assign as a vertex property
V(sf_gr)$community <- membership(communities)

# How large are they?
sizes(communities)
```

---
class: left, middle, rstudio-logo

## Louvain algorithm (2/2)

```{r, out.height = "400", out.width = "400", fig.align = "center"}
set.seed(123)
ggraph(sf_gr, layout = "fr") +
  geom_edge_link(color =  "grey") +
  geom_node_point(aes(color = as.factor(community)),
                  show.legend = FALSE) +
  theme_void()
```

---
class: left, middle, rstudio-logo, bigfont

## Assortativity

Homophily is the tendency for similar people to be connected to each other, i.e., "birds of a feather flock together". When people share common traits, it is sometimes easier to form relationships. Examples include

- education
- political beliefs
- social class
- hobbies
- age
- gender

We can measure homophily in a network using assortativity.

---
class: left, middle, rstudio-logo, bigfont

## Assortativity for nominal data

For undirected graphs:

$$
\begin{equation}
r = \frac{\sum_{i}e_{ii} - \sum_{i}a_i^2}{1-\sum_{i}a_i^2}
\end{equation}
$$

where:

- $e_{ii}$ is the fraction of edges between nodes node of type $i$ to one of type $i$.
- $a_i$ is the fraction of each type of edge that is connected to a node of type $i$

Assortativity ranges from -1 to 1, 1 meaning people only connect with people like them, 0 meaning people connect with all sorts of people equally, and -1 meaning people only connect with people unlike them.

---
class: left, middle, rstudio-logo

## Example

Suppose we want to understand if dog owners tend to be connected to other dog owners, and if non-dog owners tend to be connected to other non-dog owners.

```{r, echo = FALSE}
t <- data.frame(
  A = c(0.34, 0.2, 0.54),
  B = c(0.2, 0.26, 0.46),
  C = c(0.54, 0.46, NA)
)

row.names(t) <- c("Dog","No dog","$a_i$")
knitr::kable(t,
             row.names = TRUE,
             col.names =  c("Dog","No dog","$a_i$")) |>
    kable_minimal()

```

$$
\begin{equation}
r = \frac{(0.34+0.26)-(0.54^2+0.46^2)}{1-(0.54^2+0.46^2)} = 0.195
\end{equation}
$$

---
class: left, middle, rstudio-logo, bigfont

## Assortativity for numeric data

$$
\begin{equation}
r = \frac{\sum_{xy}xy(e_{xy}-a_xb_y)}{\sigma_a\sigma_b}
\end{equation}
$$

where:

- $e_{xy}$ is the fraction of edges joining nodes with values $x$ and $y$
- $a_x$ is the fraction of edges that start and end at nodes with values of $x$
- $b_y$ is the fraction of edges that start and end at nodes with values of $y$
- $\sigma_a$ is the standard deviation of the distribution of $a_x$
- $\sigma_b$ is the standard deviation of the distribution of $b_y$

---
class: left, middle, rstudio-logo, bigfont

## Using `igraph` to calculate assortativity

Let's use the workfrance edgelist. Suppose we want to know if people tend to work with people in their same department.

```{r}
url <- "https://ona-book.org/data/workfrance_edgelist.csv"
workfrance_edgelist <- read.csv(url)
head(workfrance_edgelist)
```

---
class: left, middle, rstudio-logo, bigfont

## Adding relevant feature

We'll need to pull in department in order to add this to the graph we will be creating.

```{r}
url <- "https://ona-book.org/data/workfrance_vertices.csv"
workfrance_vertices <- read.csv(url)
head(workfrance_vertices)
```

---
class: left, middle, rstudio-logo, bigfont

## Create graph and calculate assortativity

```{r}
gr <- workfrance_edgelist[1:2] |>
  igraph::graph_from_data_frame(
    directed = F,
    vertices = workfrance_vertices
  )

assortativity_nominal(gr, factor(V(gr)$dept))

```

---
class: left, middle, rstudio-logo

# &#127857; Lunchtime!  &#128523; 
