---
title: "Obtaining US Senator Network Data from Twitter"
author: "Keith McNulty"
date: '2022-07-08'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting the Twitter IDs of current US Senators

The Twitter IDs of US senators who are currently active on Twitter are available at the [UC San Diego Library](https://ucsd.libguides.com/congress_twitter/senators).  We can scrape these details into a data frame.

```{r}
library(rvest)
library(tidyverse)
library(urltools)

url <- "https://ucsd.libguides.com/congress_twitter/senators"
tables <- url |> 
  rvest::read_html() |> 
  rvest::html_nodes("table") 

senator_df <- tables[1] |> 
  rvest::html_table(header = TRUE) |> 
  dplyr::bind_rows(
    tables[2] |> 
      rvest::html_table(header = TRUE)
  )

senator_screen_names <- url |>
  rvest::read_html() |> 
  rvest::html_nodes("table") |> 
  rvest::html_nodes("tr") |> 
  rvest::html_node("a") |> 
  rvest::html_attr("href") |> 
  na.omit() |> 
  urltools::url_parse() |> 
  dplyr::select(screen_name = path)

senator_df <- senator_df |> 
  dplyr::bind_cols(senator_screen_names)
```

## Getting Twitter timelines

We will use the screen names to extract the last 1000 tweets for each Senator:

```{r}
library(rtweet)

results <- rtweet::get_timelines(senator_df$screen_name, n = 1000)
```

## Generating an edgelist

We use a series of functions on the `results` dataframe to generate a weighted edgelist based on interactions between Senators.   An interaction is a mention, reply, retweet or quote tweet.  The weight is a count of the interactions.

```{r}
# function to create edgelist for single MP
count_and_group <- function (df) {
  df |> 
    dplyr::select(ends_with("screen_name")) |> 
    unlist() |> 
    tibble(interacted_with = _) |> 
    tidyr::drop_na() |> 
    dplyr::group_by(interacted_with) |> 
    dplyr::summarise(weight = n()) |> 
    dplyr::filter(
      # ensures that only Senator interactions are returned
      interacted_with %in% senator_df$screen_name
    )
}

# function to generate edgelist across all MPs
create_edgelist <- function(tweet_df) {
  tweet_df |> 
    dplyr::nest_by(screen_name) |> 
    dplyr::summarise(count_and_group(data)) |> 
    # ignore interactions with self
    dplyr::filter(screen_name != interacted_with) |> 
    dplyr::rename(from = screen_name, to = interacted_with)
}

# create final edgelist
senator_edgelist <- create_edgelist(results)
```

## Creating vertex information

We gather some interesting facts and info about each active senator.

```{r}
senator_vertices <- results %>%
  select(screen_name, profile_image_url, followers_count) %>%
  distinct() |> 
  left_join(senator_df) |> 
  rename(name = Senator,
         state = State,
         party = Party)
```

Finally we ensure that there are no edges to deactivated Twitter profiles:

```{r}
senator_edgelist <- senator_edgelist |> 
  dplyr::filter(to %in% senator_vertices$screen_name)
```
